---
title: "Migration ecology of a re-established trumpeter swan population"
author: "David W. Wolfson, Randall T. Knapik, Anna Buckardt-Thomas, Laura Kearns,
  Brian Kiss, Steven Cordts, Taylor Finger, Tyler Harms, Sumner Matteson, Tiffany
  Mayo, Timothy Poole, John Moriarty, Margaret Smith, Christine Herwig, Drew Fowler,
  Thomas R. Cooper, John R. Fieberg, David E. Andersen"
output:
  pdf_document: default
  html_document: default
header-includes: \usepackage{booktabs}
---
```{r chunk-setup-options, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message=FALSE,
                      warning=FALSE,
                      fig.cap=TRUE,
                      out.width = '100%',
                      ft.keepnext=FALSE)
```

```{r flextable-settings, include=FALSE}
flextable::set_flextable_defaults(
    font.family = "New Times Roman",
    font.size = 12,
    text.align='center'
  )
```

<style>
caption {
padding-bottom: 10px;
}
</style>
<!--- this is how to write a comment that won't knit--->

<!--- --->

```{r install packages}
library(bookdown)
library(rmarkdown)
library(here)
library(knitr)
library(tidyverse)
library(flextable)
library(lubridate)
library(ggpubr)
```


## Supplemental Methods Description

Rule-based thresholds chosen:  
- Minimum distance between 2 potential segments: **2 kilometers**  
- Minimum time difference between 2 potential change points: **2 days**  
- Minimum distance moved between breeding/capture location and the furthest segment in order to consider onset of fall migration: **100 kilometers**  
- Minimum distance moved between the furthest segment and the segment representing return to spring territory: **100 kilometers**  
- Maximum distance between the spring return segment and the breeding/capture territory in order to consider a spring migration arrival: **10 kilometers**  
- Latest date to be considered a fall migration onset / earliest date to be considered a spring return onset: **1 December**  


## Supplemental Figures

```{r deployment-fig, fig.align="left", fig.cap="Figure S1. An overview of GPS telemetry data received from all collared IP trumpeter swans. Each line represents the period of data collection from a single collared swan. The grey regions indicate periods of collar deployment. The black lines are 1 January of each year. Number of deployments (including redeployments) by state/province are: Michigan (n=14), Minnesota (n=56), Manitoba (n=11), Iowa (n=12), Ohio (n=20), Wisconsin (n=9), and Arkansas (n=4)."}

# package names
packages<-c("tidyverse", "here", "lubridate", "scales", "data.table", "ggpubr")

# load packages
invisible(lapply(packages, library, character.only = TRUE))
options(scipen = 999)

# use the most recent dataset (pulled 4/28/2023)
# df<-fread(here("data/full_dataset_4_28_2023/full_w_nsd.csv"))
# 
# # julian date
# df$yday<-yday(df$timestamp)
# 
# # concatenate year and julian date
# df$jdate<-as.Date(paste(as.character(df$year), as.character(df$yday), sep="-"), "%Y-%j")
# 
# # change capture state to factor
# df$capture_state<-as.factor(df$capture_state)
# 
# # remove 1 outlier date
# df<-df %>% 
#   filter(year<2024)
# 
# sub<-df %>% distinct(jdate, id, capture_state)
# sub<-sub %>% 
#   group_by(id) %>% 
#   mutate(days=n())
# 
# sub$id<-factor(sub$id, levels=sort(unique(sub$id[order(sub$days)])))

# save out dataset to directly import in later runs
#write_csv(sub, here("data/full_dataset_4_28_2023/consolidated_julian_date.csv"))

# import consolidated dataset to save time rendering
sub<-read_csv(here("data/full_dataset_4_28_2023/consolidated_julian_date.csv"))

rect1<-data.frame(xstart=as.Date("2019-07-25"), xend=as.Date("2019-08-29"))
rect2<-data.frame(xstart=as.Date("2020-06-23"), xend=as.Date("2020-09-08"))
rect3<-data.frame(xstart=as.Date("2021-07-06"), xend=as.Date("2021-08-22"))
rect4<-data.frame(xstart=as.Date("2021-12-15"), xend=as.Date("2021-12-30"))

ggplot(sub, aes(jdate, fct_reorder(id, days, .desc=F), color=capture_state))+
  theme_pubr()+
  geom_point(size=1.5)+
  theme(axis.text.y=element_blank())+
  theme(axis.text.x=element_text(size=12, face="bold"))+
  facet_grid(fct_relevel(capture_state, 'MI','MN', 'MB', 'IA', 'OH', 'WI', 'AR')~., scales="free")+
  geom_rect(data=rect1, aes(xmin=xstart, xmax=xend, ymin=-Inf, ymax=Inf,
  ), alpha=0.3, inherit.aes=F)+
  geom_rect(data=rect2, aes(xmin=xstart, xmax=xend, ymin=-Inf, ymax=Inf,
  ), alpha=0.3, inherit.aes=F)+
  geom_rect(data=rect3, aes(xmin=xstart, xmax=xend, ymin=-Inf, ymax=Inf,
  ), alpha=0.3, inherit.aes=F)+
  geom_rect(data=rect4, aes(xmin=xstart, xmax=xend, ymin=-Inf, ymax=Inf,
  ), alpha=0.3, inherit.aes=F)+
  theme(axis.title.y=element_blank())+
  theme(axis.title.x=element_blank())+
  theme(strip.text.y = element_text(size=12, face="bold"))+
  theme(strip.background = element_rect(color="black", fill="white"))+
  guides(colour="none", fill="none")+
  scale_x_date(date_breaks = "1 year", date_labels = "%b-%Y")+
  geom_vline(xintercept = as.numeric(ymd("2020-01-01")), color="black", linewidth=1)+
  geom_vline(xintercept = as.numeric(ymd("2021-01-01")), color="black", linewidth=1)+
  geom_vline(xintercept = as.numeric(ymd("2022-01-01")), color="black", linewidth=1)+
  geom_vline(xintercept = as.numeric(ymd("2023-01-01")), color="black", linewidth=1)+
  theme(panel.spacing.y = unit(0, "lines"))
```



```{r mig-extent-vs-latitude-fig, fig.align="left", fig.cap="Figure S2. Breeding/capture latitude versus extent of migration (furthest distance from breeding territory during the nonbreeding season) for 221 'swan-year' datasets representing annual migration cycles."}

df<-read_csv(here("data/full_dataset_4_28_2023/daily_nsd.csv"))
# 126 separate collar deployments

# split years up each summer
df <- df %>%
  group_by(id) %>%
  mutate(swan_yr = ifelse(yday < 182, paste(id, year - 1, year, sep = "-"),
                          paste(id, year, year + 1, sep = "-")
  )) # 182 is julian day for july 1

df<-df %>% 
  group_by(swan_yr) %>% 
  mutate(num_days=n())

# filter out swans that had years with less than 90 days
df<-df %>% 
  filter(num_days>90)

# filtered out 21 swan-years with less than 90 days; 231 swan-year combinations

# Convert NSD to simple displacement in kilometers
df$sqrt<-sqrt(df$nsd_daily_mean)
df$rescale<-df$sqrt/1000

#########################################################################

# Pull out the maximum displacement value for each swan-year
df<-df %>% 
  group_by(swan_yr) %>% 
  mutate(max_nsd=max(rescale)) %>% 
  select(id, capture_state, state_ID, sex, swan_yr, max_nsd, num_days) %>% 
  distinct()


# arkansas data separate from the overall mcp fit
# ark<-read_csv(here("output/arkansas_migration_metrics_salvaged.csv"))
# The numbers that I pulled off from AR swans actually agrees pretty closely with 
# what mcp got from the batch run, so I'll leave it for now (8/18/2023).

# others to exclude
exclude<-c(
  "1P-2020-2021", # taken into custody, year all screwy
  "9J (swan originally collared as 5J)-2021-2022", # collar died before winter
  "5L-2020-2021" ,"5L-2021-2022", # the cygnet that went up to Hudson Bay
  "6M-2021-2022", "6M-2022-2023", # Ohio disperser
  "7M-2021-2022", # Dropped 7M because it made a big movement NE into Pennsylvania (and then back to territory)
  "8P-2021-2022", # big summer dispersal N and then collar died
  "9N-2021-2022", # big summer dispersal
  "9N-2022-2023"  # big summer dispersal
)

# add in breeding lat and other info
ids<-read_csv(here("ids.csv"))
df<-df %>% 
  left_join(., ids) %>% 
  select(-comments) %>% 
  rename(breeding_status="breeding_status(if cygnets=breeder; if mate=paired;else non-breeder or cygnet)",
         mass='mass (kg)', skull='skull (mm)')

df<-df %>% 
  filter(!swan_yr%in%exclude)
# excluded 10 more swan-year datasets, from 231 to 221


dist_plot<-df %>% 
  ggplot(aes(breeding_lat,max_nsd))+
  geom_point()+
  # geom_hline(yintercept = quantile(p_dates$mig_extent, 0.5, na.rm=T), 
  #            lty=2, color="red")+
  # geom_text(aes(52, quantile(p_dates$mig_extent, 0.5, na.rm=T),
  #               label="50% quantile", vjust=-1), size=5, color="red")+
  # geom_hline(yintercept = quantile(p_dates$mig_extent, 0.25, na.rm=T), 
  #            lty=2, color="blue")+
  #  geom_text(aes(52, quantile(p_dates$mig_extent, 0.25, na.rm=T),
  #               label="25% quantile", vjust=-1), size=5, color="blue")+
  # geom_hline(yintercept = quantile(p_dates$mig_extent, 0.75, na.rm=T), 
  #            lty=2, color="darkgreen")+
  #  geom_text(aes(52, quantile(p_dates$mig_extent, 0.75, na.rm=T),
#               label="75% quantile", vjust=-1), size=5, color="darkgreen")+
labs(x="\nBreeding/Capture Latitude",
     y="Extent of Migration (km)\n")+
  theme_pubr()

dist_plot
```


## Migration Phenology Summary Statistics:

### Autumn Departure

We estimated migration duration for all swans that traveled >100km from the breeding/capture territory by 1 December. 

```{r combined-fall-table, ft.align="left"}
# third round (post apr/may 2023)
p_dates<-read_csv(here("output/metrics_3rd_round_manuscript_ready.csv"))

# first, show a table with summary stats for fall departure and spring arrival
# all swans and years, fall departures
fall_table<-p_dates %>% 
  drop_na(fall_mig_onset) %>% 
  summarize(num_swans=length(unique(swan_ID)),
            num_swan_years=n(),
            average_fall_depart=format(mean(as.POSIXct(fall_mig_onset)), "%B %d"),
            stan_dev_fall_depart=round(sd(fall_mig_onset)),
            earliest_fall_depart=format(range(fall_mig_onset)[1], "%B %d"),
            latest_fall_depart=format(range(fall_mig_onset)[2], "%B %d"))

# retain the total number of tracked swans (not just those that fit the rules for fall departure)
total_tracked<-data.frame(total_tracked=length(unique(p_dates$swan_ID)))

fall_table<-cbind.data.frame(total_tracked, fall_table)

fall_table %>% 
  flextable() %>% 
  set_header_labels(values=list(
    "total_tracked"="Total Swans Tracked",
    "num_swans" = "Number of Long-Distance Migrants",
    "num_swan_years" = "Number of Fall Departure Events",
    "average_fall_depart" = "Average Fall Departure",
    "stan_dev_fall_depart" = "Standard Deviation (in days)",
    "earliest_fall_depart" = "Earliest Departure",
    "latest_fall_depart" = "Latest Departure")) %>% 
  set_table_properties(layout="autofit") %>% 
   fontsize(size=12, part="all") %>% 
  vline(part="header") %>% 
  align(align ='center', part = 'all') %>% 
  bg(part="header", bg="#dbdbdb") %>% 
  bold(part="header") %>% 
  set_caption("Table S1. Compiled migration phenology of all fall departures from 2019-2022.")
```


```{r fall-yearly-table, ft.align="left"}
# third round (post apr/may 2023)
p_dates<-read_csv(here("output/metrics_3rd_round_manuscript_ready.csv"))

# Split out by year
fall_yearly<-p_dates %>% 
  drop_na(fall_mig_onset) %>% 
  # filter(!fall_yr==2019) %>% 
  group_by(fall_yr) %>% 
  summarize(num_swans=length(unique(swan_ID)),
            average_fall_depart=format(mean(as.POSIXct(fall_mig_onset)), "%B %d"),
            stan_dev_fall_depart=round(sd(fall_mig_onset)),
            earliest_fall_depart=format(range(fall_mig_onset)[1], "%B %d"),
            latest_fall_depart=format(range(fall_mig_onset)[2], "%B %d"))

# retain the total number of tracked swans (not just those that fit the rules for fall departure)
tracked_by_year<-p_dates %>% 
  group_by(fall_yr) %>% 
  summarize(tracked_swans=length(unique(swan_ID)))

fall_yearly<-cbind.data.frame(tracked_by_year, fall_yearly[,2:6])

fall_yearly %>% 
  flextable() %>% 
  colformat_num(big.mark="") %>% 
  set_header_labels(values=list(
    "tracked_swans"="Total Swans Tracked",
    "fall_yr" = "Year",
    "num_swans" = "Number of Long-Distance Migrants",
    "average_fall_depart" = "Average Fall Departure",
    "stan_dev_fall_depart" = "Standard Deviation (in days)",
    "earliest_fall_depart" = "Earliest Departure",
    "latest_fall_depart" = "Latest Departure")) %>% 
      set_table_properties(layout="autofit") %>% 
  vline(part="header") %>% 
  align(align ='center', part = 'all') %>% 
  bg(part="header", bg="#dbdbdb") %>% 
  bold(part="header") %>% 
  set_caption("Table S2. Yearly summaries of migration phenology of fall departures from 2019-2022.")
```


```{r fall-breeding-table, ft.align="left"}
# third round (post apr/may 2023)
p_dates<-read_csv(here("output/metrics_3rd_round_manuscript_ready.csv"))

# retain the total number of tracked swans (not just those that fit the rules for fall departure)
tracked_by_breeding<-p_dates %>% 
  drop_na(breeding_status) %>% 
  filter(breeding_status%in%c("breeder", "non_breeder", "paired")) %>%    #drop cygnets
  mutate(breeding_status=recode(.$breeding_status, breeder="Breeder", non_breeder="Non-Breeder", paired="Paired")) %>% 
  group_by(breeding_status) %>% 
  summarize(tracked_swans=length(unique(swan_ID)))

# Split by breeding status
# Fall
fall_breeding<-p_dates %>% 
  drop_na(fall_mig_onset) %>% 
  drop_na(breeding_status) %>% 
  filter(breeding_status%in%c("breeder", "non_breeder", "paired")) %>%
  group_by(breeding_status) %>% 
  summarize(num_swans=length(unique(swan_ID)),
            num_swan_years=n(),
            average_fall_depart=format(mean(as.POSIXct(fall_mig_onset)), "%B %d"),
            # median_fall_depart=format(median(as.POSIXct(fall_mig_onset)), "%B %d"),
            stan_dev_fall_depart=round(sd(fall_mig_onset)),
            earliest_fall_depart=format(range(fall_mig_onset)[1], "%B %d"),
            latest_fall_depart=format(range(fall_mig_onset)[2], "%B %d"))

fall_breeding<-cbind.data.frame(tracked_by_breeding, fall_breeding[,2:7])


fall_breeding %>% 
  flextable() %>% 
  set_header_labels(values=list(
    "breeding_status" = "Breeding Status",
    "tracked_swans" = "Total Swans Tracked",
    "num_swans" = "Number of Long-Distance Migrants",
    "num_swan_years" = "Number of Fall Departure Events",
    "average_fall_depart" = "Average Fall Departure",
    "stan_dev_fall_depart" = "Standard Deviation (in days)",
    "earliest_fall_depart" = "Earliest Departure",
    "latest_fall_depart" = "Latest Departure")) %>% 
  set_table_properties(layout="autofit") %>% 
  vline(part="header") %>% 
  align(align ='center', part = 'all') %>% 
  bg(part="header", bg="#dbdbdb") %>% 
  bold(part="header") %>% 
  set_caption("Table S3. Fall departures of long-distance migrants by breeding status")
```


### Spring Arrival

We estimated spring arrival for all swans that traveled >100km from the breeding/capture territory during the non-breeding season and then returned to within 10 km of their previous summer territory.  

```{r combined-spring-table, ft.align="left"}
# third round (post apr/may 2023)
p_dates<-read_csv(here("output/metrics_3rd_round_manuscript_ready.csv"))

# all swans and years, spring arrivals
spring_table<-p_dates %>% 
  drop_na(spring_arrival) %>% 
  summarize(num_swans=length(unique(swan_ID)),
            num_swan_years=n(),
            average_spring_arrival=format(mean(as.POSIXct(spring_arrival)), "%B %d"),
            stan_dev_spring_arrival=round(sd(spring_arrival)),
            earliest_spring_arrival=format(range(spring_arrival)[1], "%B %d"),
            latest_spring_arrival=format(range(spring_arrival)[2], "%B %d"))

# retain the total number of tracked swans (not just those that fit the rules for fall departure)
total_tracked<-data.frame(total_tracked=length(unique(p_dates$swan_ID)))

spring_table<-cbind.data.frame(total_tracked, spring_table)

spring_table %>% 
  flextable() %>% 
  set_header_labels(values=list(
    "total_tracked"="Total Swans Tracked",
    "num_swans" = "Number of Long-Distance Migrants",
    "num_swan_years" = "Number of Spring Arrival Events",
    "average_spring_arrival" = "Average Spring Arrival",
    "stan_dev_spring_arrival" = "Standard Deviation (in days)",
    "earliest_spring_arrival" = "Earliest Arrival",
    "latest_spring_arrival" = "Latest Arrival")) %>% 
  set_table_properties(layout="autofit") %>% 
  vline(part="header") %>% 
  align(align ='center', part = 'all') %>% 
  bg(part="header", bg="#dbdbdb") %>% 
  bold(part="header") %>% 
  set_caption("Table S4. Compiled migration phenology of all spring arrivals from 2020-2023.")
```



```{r spring-yearly-table, ft.align="left"}
# third round (post apr/may 2023)
p_dates<-read_csv(here("output/metrics_3rd_round_manuscript_ready.csv"))

spring_yearly<-p_dates %>% 
  drop_na(spring_arrival) %>% 
  group_by(spring_yr) %>% 
  summarize(num_swans=length(unique(swan_ID)),
            average_spring_arrival=format(mean(as.POSIXct(spring_arrival)), "%B %d"),
            stan_dev_spring_arrival=round(sd(spring_arrival)),
            earliest_spring_arrival=format(range(spring_arrival)[1], "%B %d"),
            latest_spring_arrival=format(range(spring_arrival)[2], "%B %d"))

# retain the total number of tracked swans (not just those that fit the rules for fall departure)
tracked_by_year<-p_dates %>% 
  group_by(spring_yr) %>% 
  summarize(tracked_swans=length(unique(swan_ID)))

spring_yearly<-cbind.data.frame(tracked_by_year, spring_yearly[,2:6])

spring_yearly %>% 
  flextable() %>% 
  colformat_num(big.mark="") %>% 
  set_header_labels(values=list(
    "spring_yr" = "Year",
    "tracked_swans" = "Total Swans Tracked",
    "num_swans" = "Number of Long-Distance Migrants",
    "average_spring_arrival" = "Average Spring Arrival",
    "stan_dev_spring_arrival" = "Standard Deviation (in days)",
    "earliest_spring_arrival" = "Earliest Arrival",
    "latest_spring_arrival" = "Latest Arrival")) %>% 
  set_table_properties(layout="autofit") %>% 
  vline(part="header") %>% 
  align(align ='center', part = 'all') %>% 
  bg(part="header", bg="#dbdbdb") %>% 
  bold(part="header") %>% 
  set_caption("Table S5. Yearly summaries of migration phenology of spring arrivals from 2020-2023.")
```




```{r spring-breeding-table, ft.align="left"}
# third round (post apr/may 2023)
p_dates<-read_csv(here("output/metrics_3rd_round_manuscript_ready.csv"))

# retain the total number of tracked swans (not just those that fit the rules for fall departure)
tracked_by_breeding<-p_dates %>% 
  drop_na(breeding_status) %>% 
  filter(breeding_status%in%c("breeder", "non_breeder", "paired")) %>%    #drop cygnets
  mutate(breeding_status=recode(.$breeding_status, breeder="Breeder", non_breeder="Non-Breeder", paired="Paired")) %>% 
  group_by(breeding_status) %>% 
  summarize(tracked_swans=length(unique(swan_ID)))

# Split by breeding status
# Spring
spring_breeding<-p_dates %>% 
  drop_na(spring_arrival) %>% 
  drop_na(breeding_status) %>% 
  filter(breeding_status%in%c("breeder", "non_breeder", "paired")) %>%
  group_by(breeding_status) %>% 
  summarize(num_swans=length(unique(swan_ID)),
            num_swan_years=n(),
            average_spring_arrival=format(mean(as.POSIXct(spring_arrival)), "%B %d"),
            stan_dev_spring_arrival=round(sd(spring_arrival)),
            earliest_spring_arrival=format(range(spring_arrival)[1], "%B %d"),
            latest_spring_arrival=format(range(spring_arrival)[2], "%B %d"))



spring_breeding<-cbind.data.frame(tracked_by_breeding, spring_breeding[,2:7])

spring_breeding %>% 
  flextable() %>% 
  set_header_labels(values=list(
    "breeding_status" = "Breeding Status",
    "tracked_swans" = "Total Swans Tracked",
    "num_swans" = "Number of Long-Distance Migrants",
    "num_swan_years" = "Number of Spring Arrival Events",
    "average_spring_arrival" = "Average Spring Arrival",
    "stan_dev_spring_arrival" = "Standard Deviation (in days)",
    "earliest_spring_arrival" = "Earliest Arrival",
    "latest_spring_arrival" = "Latest Arrival")) %>%  
  set_table_properties(layout="autofit") %>% 
  vline(part="header") %>% 
  align(align ='center', part = 'all') %>% 
  bg(part="header", bg="#dbdbdb") %>% 
  bold(part="header") %>% 
  set_caption("Table S6. Spring arrivals of long-distance migrants by breeding status") %>% 
  width(., width = dim(.)$widths * 6.5 / (flextable::flextable_dim(.)$widths))
```



### Migration Duration

We estimated migration duration for all swans that traveled >100km from the breeding/capture territory during the non-breeding season and then returned to within 10 km of their previous summer territory. Migration duration represents the span of time absent from the breeding/capture territory during the non-breeding season, and is calculated by the difference in days between spring arrival and the previous year's autumn departure. 

```{r duration-pooled-table, ft.align="left"}
# third round (post apr/may 2023)
p_dates<-read_csv(here("output/metrics_3rd_round_manuscript_ready.csv"))

# all swans and years, migration duration
duration_table<-p_dates %>% 
  drop_na(mig_duration) %>% 
  summarize(num_swans=length(unique(swan_ID)),
            num_swan_years=n(),
            average_migration_duration=round(mean(mig_duration)),
            stan_duration=round(sd(mig_duration)))
            # shortest_duration=round(min(mig_duration)),
            # longest_duration=round(max(mig_duration)))

# retain the total number of tracked swans (not just those that fit the rules for fall departure)
total_tracked<-data.frame(total_tracked=length(unique(p_dates$swan_ID)))

duration_table<-cbind.data.frame(total_tracked, duration_table)

duration_table%>% 
  flextable() %>% 
  set_header_labels(values=list(
    "total_tracked"="Total Swans Tracked",
    "num_swans" = "Number of Long-Distance Migrants",
    "num_swan_years" = "Number of Annual Cycles",
    "average_migration_duration" = "Average Migration Duration",
    "stan_duration" = "Standard Deviation (in days)")) %>% 
    # "shortest_duration" = "Shortest Duration",
    # "longest_duration" = "Longest Duration")) %>% 
  set_table_properties(layout="autofit") %>% 
  vline(part="header") %>% 
  align(align ='center', part = 'all') %>% 
  bg(part="header", bg="#dbdbdb") %>% 
  bold(part="header") %>% 
  set_caption("Table S7. Compiled migration duration for all swans from 2019-2023.")
```

```{r departure-yearly-table, ft.align="left"}
# third round (post apr/may 2023)
p_dates<-read_csv(here("output/metrics_3rd_round_manuscript_ready.csv"))

# all swans and years, migration duration
duration_table<-p_dates %>% 
  drop_na(mig_duration) %>% 
   group_by(spring_yr) %>% 
  summarize(num_swans=length(unique(swan_ID)),
            average_migration_duration=round(mean(mig_duration)),
            stan_duration=round(sd(mig_duration)))
            # shortest_duration=round(min(mig_duration)),
            # longest_duration=round(max(mig_duration)))

# retain the total number of tracked swans (not just those that fit the rules for fall departure)
tracked_by_year<-p_dates %>% 
  group_by(spring_yr) %>% 
  summarize(tracked_swans=length(unique(swan_ID)))

duration_table<-cbind.data.frame(tracked_by_year, duration_table[,2:4])

duration_table%>% 
  flextable() %>% 
  colformat_num(big.mark="") %>% 
  set_header_labels(values=list(
    "spring_yr" = "Year",
    "tracked_swans"="Total Swans Tracked",
    "num_swans" = "Number of Long-Distance Migrants",
    "average_migration_duration" = "Average Migration Duration",
    "stan_duration" = "Standard Deviation (in days)")) %>% 
    # "shortest_duration" = "Shortest Duration",
    # "longest_duration" = "Longest Duration")) %>% 
  set_table_properties(layout="autofit") %>% 
  vline(part="header") %>% 
  align(align ='center', part = 'all') %>% 
  bg(part="header", bg="#dbdbdb") %>% 
  bold(part="header") %>% 
  set_caption("Table S8. Yearly summaries of migration duration from 2019-2020 until 2022-2023.")
```


```{r duration-breeding-table, ft.align="left"}
# third round (post apr/may 2023)
p_dates<-read_csv(here("output/metrics_3rd_round_manuscript_ready.csv"))

# all swans and years, migration duration
duration_table<-p_dates %>% 
  drop_na(mig_duration) %>% 
  group_by(breeding_status) %>% 
  summarize(num_swans=length(unique(swan_ID)),
            average_migration_duration=round(mean(mig_duration)),
            stan_duration=round(sd(mig_duration)))
            #shortest_duration=round(min(mig_duration)),
            #longest_duration=round(max(mig_duration)))

# retain the total number of tracked swans (not just those that fit the rules for fall departure)
tracked_by_breeding<-p_dates %>% 
  drop_na(breeding_status) %>% 
  filter(breeding_status%in%c("breeder", "non_breeder", "paired")) %>%    #drop cygnets
  mutate(breeding_status=recode(.$breeding_status, breeder="Breeder", non_breeder="Non-Breeder", paired="Paired")) %>%
  group_by(breeding_status) %>% 
  summarize(tracked_swans=length(unique(swan_ID)))

duration_table<-cbind.data.frame(tracked_by_breeding, duration_table[,2:4])

duration_table%>% 
  flextable() %>% 
  colformat_num(big.mark="") %>% 
  set_header_labels(values=list(
        "breeding_status" = "Breeding Status",
    "tracked_swans"="Total Swans Tracked",
    "num_swans" = "Number of Long-Distance Migrants",
    "average_migration_duration" = "Average Migration Duration",
    "stan_duration" = "Standard Deviation (in days)")) %>% 
    #"shortest_duration" = "Shortest Duration",
    #"longest_duration" = "Longest Duration")) %>% 
  set_table_properties(layout="autofit") %>% 
  vline(part="header") %>% 
  align(align ='center', part = 'all') %>% 
  bg(part="header", bg="#dbdbdb") %>% 
  bold(part="header") %>% 
  set_caption("Table S9. Summaries by breeding status of migration duration from 2019-2020 until 2022-2023.")
```



